The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2018-04-23 script/en/chapters/primes.tex
CT Book ed 12: Deleted some comments; added Python to Index.
CT1: Updated CT book (d+e): typos, dead links, 2018, 50th M-prime, automated rowno in 2 tables in chap 3, Make.rules also for Windows

\chapter{Prime Numbers}
\label{Chapter_Primes}
(\hyperlink{author_Bernhard-Esslinger}{Bernhard Esslinger}, May 1999;
 Updates: Nov 2000, Dec 2001, Jun 2003, May 2005, Mar 2006, Jun 2007,
          Jan 2010, Aug 2013, Jul 2016)

@@

% --------------------------------------------------------------------------
% --------------------------------------------------------------------------
\vskip + 20pt
\section{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 3.4

The largest prime numbers known today have several millions digits, 
which is too big for us to imagine. The number of elementary particles in
the universe is estimated to be ``only'' a $80$-digit decimal number 
\hyperlink{grosord}{(see the overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 30+ largest known primes (as of June 2016)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet
at   \url{https://primes.utm.edu/primes/search.php?Number=1000},
at   \url{http://primes.utm.edu/mersenne/index.html}, and
at   \url{http://www.mersenne.org/primes/}.
}:

\index{Mersenne!number!generalized}
\index{Fermat!number!generalized} 
\begin{table}[ht] % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt}    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline \rule{0pt}{15pt}
	1  & $2^{74,207,281}-1$ & 22,338,618 & 2016 & Mersenne, 49th known \\
	2  & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	3  & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, 47th known \\
	4  & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, 46th known \\
	5  & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, 45th known \\
	6  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, M-44 \\
	7  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, M-43 \\
	8  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	9  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	10 & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	11 & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	12 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\%%%

	13 & $3 \cdot 2^{11,895,718}-1$ & 3,580,969 & 2015 & Generalized Mersenne \\
	14 & $3 \cdot 2^{11,731,850}-1$ & 3,531,640 & 2015 & Generalized Mersenne \\
	15 & $3 \cdot 2^{11,484,018}-1$ & 3,457,035 & 2014 & Generalized Mersenne \\
	16 & $3 \cdot 2^{10,829,346}+1$ & 3,259,959 & 2014 & Generalized Mersenne \\
	17 & $ 475,856^{524,288}+1$ &   2,976,633 & 2012 & Generalized Fermat \\
	18 & $ 356,926^{524,288}+1$ &   2,911,151 & 2012 & Generalized Fermat \\
	19 & $ 341,112^{524,288}+1$ &   2,900,832 & 2012 & Generalized Fermat \\

	20 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\%%%

	21 & $90,527 \cdot 2^{9,162,167}+1$ & 2,758,093 & 2010 & Generalized Mersenne \\
	22 & $2,038 \cdot 366^{1,028,507}-1$ & 2,636,562 & 2016 & Generalized Fermat \\
	23 & $75,898^{524,288}+1$ & 2,558,647 & 2011 & Generalized Fermat \\

	24 & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\%%%

	25 & $502,573 \cdot 2^{7,181,987}-1$ & 2,162,000 & 2014 & Generalized Mersenne \\
	26 & $402,539 \cdot 2^{7,173,024}-1$ & 2,159,301 & 2014 & Generalized Mersenne \\
	27 & $161,041 \cdot 2^{7,107,964}+1$ & 2,139,716 & 2015 & Generalized Mersenne \\
	28 & $3 \cdot 2^{7,033,641}+1$ & 2,117,338 & 2011 & Generalized Mersenne \\
	29 & $33,661 \cdot 2^{7,031,232}+1$ & 2,116,617 & 2007 & Generalized Mersenne \\

	30  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
        ... &  &  &  & \\
%	16  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
%	17  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
%	18  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
%	19  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
%	20  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
%	21  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
	328 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
%	23  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
%	24  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
        ... &  &  &  & \\
	342  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 30+ largest known primes and its particular number types
         (as of June 2016)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB)  (\url{https://en.wikipedia.org/wiki/Seventeen_or_Bust})
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

@@

The larges currently known prime is a Mersenne prime, found by the
\hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}).
Within the largest known primes there are also numbers of the type
\hyperlink{generalizedMersennenumbers}{generalized Mersenne number}
(chapter~\ref{generalized-mersenne-no1})
and 
\hyperlink{generalizedFermatprimes}{generalized Fermat numbers}
(chapter~\ref{generalized-fermat}).

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known $49$
 % Eyecatcher_neue-Mersenne     % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers\index{prime number!Mersenne}\index{Mersenne!prime number}
have the following exponents:\footnote{%
  Landon Curt Noll\index{Noll, Landon Curt} lists all known Mersenne primes
  including its date of discovery and its value as number and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}.
  Also see:
      \url{http://www.utm.edu/}.
} 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609, ~ 57,885,161, ~ 74,207,281.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$

@@

\vskip +10 pt
For the first 44  % be_2005_UPDATEN_if-new-mersenne-prime-appears  % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 49th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely.\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/primenet/}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{prime number!test}. }

% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
As of 2016-07-19 all prime exponents smaller than $ 36,058,573 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
  }:
So we can
be certain, that this is really the 41th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime). Here are some examples in more detail:

@@

Right now the GIMPS project has discovered 15
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|}
\hline \rule{0pt}{10pt} 
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline \rule{0pt}{15pt}

	$2^{74,207,281}-1$ & 22,338,618 & January 7, 2016 & Curtis Cooper  \\
	$2^{57,885,161}-1$ & 17,425,170 & January 25, 2013 & Curtis Cooper  \\
	$2^{43,112,609}-1$ & 12,978,189 & August 23, 2008 & Edson Smith  \\
	$2^{42,643,801}-1$ & 12,837,064 & April 12, 2009 & Odd Magnar Strindmo \\
	$2^{37,156,667}-1$ & 11,185,272 & September 6, 2008 & Hans-Michael Elvenich \\

	$2^{32,582,657}-1$ &  9,808,358 & September 4, 2006 & Curtis Cooper/Steven Boone \\
	$2^{30,402,457}-1$ &  9,152,052 & December 15, 2005 & Curtis Cooper/Steven Boone \\
	$2^{25,964,951}-1$ &  7,816,230 & February 18, 2005 & Martin Nowak     \\
	$2^{24,036,583}-1$ &  7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ &  6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ &  4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ &  2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &    909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &    895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &    420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of January 2016)}
\label{Primes_GIMPS-Primes_table-reference}
\end{center}
\end{table}

%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

@@

\vskip +25 pt
\paragraph*{Strong pseudo prime numbers}%
\index{prime number!strong pseudo prime}\index{number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
  In 1976 an efficient probabilistic primality test was published by Prof. Rabin,
  based on a number theoretic result of Prof. Miller from the year before.\\
  Prof. Rabin worked at the Harvard and Hebrew University.
  Michael Oser Rabin was born 1931 in Breslau. His family emigrated in
  1935 to Palestine because of the Nazi politics.
  \\
  Prof. Miller worked at the Carnegie-Mellon University, School of Computer
  Science.\\
  The functionality of the  Miller-Rabin test is explained in detail in
  \cite{Witten2010b} using a Python program.
}%
: It is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

>>>>

\chapter{Prime Numbers}
\label{Chapter_Primes}
(\hyperlink{author_Bernhard-Esslinger}{Bernhard Esslinger}, May 1999;
 Updates: Nov 2000, Dec 2001, Jun 2003, May 2005, Mar 2006, Jun 2007,
          Jan 2010, Aug 2013, Jul 2016, Apr 2018)

@@

% --------------------------------------------------------------------------
% --------------------------------------------------------------------------
%% \vskip + 20pt
\newpage
\section{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 3.4

The largest prime numbers known today have several millions digits.\footnote{%
	In CT1\index{CT1} you can calculate all digits of such a big number
	very quickly via the menu path {\bf Indiv. Procedures \textbackslash{}
	Number Theory Interactive \textbackslash{} Compute Mersenne Numbers}.
} 
This is too big for us to imagine. The number of elementary particles in
the universe is estimated to be ``only'' a $80$-digit decimal number 
\hyperlink{grosord}{(see the overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 30+ largest known primes (as of Jan 2018)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table \ref{L_n_Largest_Known-Primes} contains the biggest,
currently known primes and a description of its particular number
type.\footnote{%
An up-to-date version can be found in the internet
at   \url{https://primes.utm.edu/primes/search.php?Number=1000},
at   \url{http://primes.utm.edu/mersenne/index.html}, and
at   \url{http://www.mersenne.org/primes/}.
}

\index{Mersenne!number!generalized}
\index{Fermat!number!generalized}

\ignoreoutput{\rowno[0]}
\begin{table}[ht] % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt}    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline \rule{0pt}{15pt}
	\rowno & $2^{77,232,917}-1$ & 23,249,425 & 2018 & Mersenne, 50th known \\
	\rowno & $2^{74,207,281}-1$ & 22,338,618 & 2016 & Mersenne, 49th known \\
	\rowno & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	\rowno & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, M-47 \\
	\rowno & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, M-46 \\
	\rowno & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, M-45 \\
	\rowno & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, M-44 \\
	\rowno & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, M-43 \\
	\rowno & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	\rowno & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	\rowno & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	\rowno & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	\rowno & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\%%%

	\rowno & $3 \cdot 2^{11,895,718}-1$ & 3,580,969 & 2015 & Generalized Mersenne \\
	\rowno & $3 \cdot 2^{11,731,850}-1$ & 3,531,640 & 2015 & Generalized Mersenne \\
	\rowno & $3 \cdot 2^{11,484,018}-1$ & 3,457,035 & 2014 & Generalized Mersenne \\
	\rowno & $3 \cdot 2^{10,829,346}+1$ & 3,259,959 & 2014 & Generalized Mersenne \\
	\rowno & $ 475,856^{524,288}+1$ &   2,976,633 & 2012 & Generalized Fermat \\
	\rowno & $ 356,926^{524,288}+1$ &   2,911,151 & 2012 & Generalized Fermat \\
	\rowno & $ 341,112^{524,288}+1$ &   2,900,832 & 2012 & Generalized Fermat \\

	\rowno & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\%%%

	\rowno & $90,527 \cdot 2^{9,162,167}+1$ & 2,758,093 & 2010 & Generalized Mersenne \\
	\rowno & $2,038 \cdot 366^{1,028,507}-1$ & 2,636,562 & 2016 & Generalized Fermat \\
	\rowno & $75,898^{524,288}+1$ & 2,558,647 & 2011 & Generalized Fermat \\

	\rowno & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\%%%

	\rowno & $502,573 \cdot 2^{7,181,987}-1$ & 2,162,000 & 2014 & Generalized Mersenne \\
	\rowno & $402,539 \cdot 2^{7,173,024}-1$ & 2,159,301 & 2014 & Generalized Mersenne \\
	\rowno & $161,041 \cdot 2^{7,107,964}+1$ & 2,139,716 & 2015 & Generalized Mersenne \\
	\rowno & $3 \cdot 2^{7,033,641}+1$ & 2,117,338 & 2011 & Generalized Mersenne \\
	\rowno & $33,661 \cdot 2^{7,031,232}+1$ & 2,116,617 & 2007 & Generalized Mersenne \\

	\rowno  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
        ... &  &  &  & \\
%	16  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
%	17  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
%	18  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
%	19  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
%	20  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
%	21  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
        % Eyecatcher_neue-Mersenne
	329 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
%	23  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
%	24  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
        ... &  &  &  & \\
        % Eyecatcher_neue-Mersenne
	343  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 30+ largest known primes and its particular number types
         (as of Jan 2018)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB)  (\url{https://en.wikipedia.org/wiki/Seventeen_or_Bust})
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

\clearpage

@@

The largest currently known prime is a Mersenne prime, found by the
\hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}).
Within the largest known primes there are also numbers of the type
\hyperlink{generalizedMersennenumbers}{generalized Mersenne number}
(chapter~\ref{generalized-mersenne-no1})
and 
\hyperlink{generalizedFermatprimes}{generalized Fermat numbers}
(chapter~\ref{generalized-fermat}).

\clearpage

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known $50$
 % Eyecatcher_neue-Mersenne     % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers\index{prime number!Mersenne}\index{Mersenne!prime number}
have the following exponents:\footnote{%
  Landon Curt Noll\index{Noll, Landon Curt} lists all known Mersenne primes
  including its date of discovery and its value as number and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}.
  Also see:
      \url{http://www.utm.edu/}.
} 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609, ~ 57,885,161, ~ 74,207,281, ~ 77,232,917.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$

@@

\vskip +10 pt
For the first 47  % be_2005_UPDATEN_if-new-mersenne-prime-appears  % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 50th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely.\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/primenet/}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{prime number!test}. }

% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
As of 2018-04-23 all prime exponents smaller than $ 43,261,403 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
  }:
% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
So we can be certain, that this is really the 47th
Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime). Here are some examples in more detail:

@@

Right now the GIMPS project has discovered 16 %% https://www.mersenne.org/primes/
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\ignoreoutput{\rowno[0]}
\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt} 
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline \rule{0pt}{15pt}

	\rowno & $2^{77,232,917}-1$ & 23,249,425 & Dec 26, 2017 & Jonathan Pace  \\
	\rowno & $2^{74,207,281}-1$ & 22,338,618 & Jan 7, 2016 & Curtis Cooper  \\
	\rowno & $2^{57,885,161}-1$ & 17,425,170 & Jan 25, 2013 & Curtis Cooper  \\
	\rowno & $2^{43,112,609}-1$ & 12,978,189 & Aug 23, 2008 & Edson Smith  \\
	\rowno & $2^{42,643,801}-1$ & 12,837,064 & Apr 12, 2009 & Odd Magnar Strindmo \\
	\rowno & $2^{37,156,667}-1$ & 11,185,272 & Sep 6, 2008 & Hans-Michael Elvenich \\

	\rowno & $2^{32,582,657}-1$ &  9,808,358 & Sep 4, 2006 & Curtis Cooper/Steven Boone \\
	\rowno & $2^{30,402,457}-1$ &  9,152,052 & Dec 15, 2005 & Curtis Cooper/Steven Boone \\
	\rowno & $2^{25,964,951}-1$ &  7,816,230 & Feb 18, 2005 & Martin Nowak     \\
	\rowno & $2^{24,036,583}-1$ &  7,235,733 & May 15, 2004 & Josh Findley     \\
	\rowno & $2^{20,996,011}-1$ &  6,320,430 & Nov 17, 2003 & Michael Shafer   \\
	\rowno & $2^{13,466,917}-1$ &  4,053,946 & Nov 14, 2001 & Michael Cameron  \\
	\rowno & $2^{ 6,972,593}-1$ &  2,098,960 & Jun 1, 1999 & Nayan Hajratwala \\
	\rowno & $2^{ 3,021,377}-1$ &    909,526 & Jan 27, 1998 & Roland Clarkson  \\
	\rowno & $2^{ 2,976,221}-1$ &    895,932 & Aug 24, 1997 & Gordon Spence    \\
	\rowno & $2^{ 1,398,269}-1$ &    420,921 & November 1996 & Joel Armengaud   \\

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of January 2018)}
\label{Primes_GIMPS-Primes_table-reference}
\end{center}
\end{table}

%%% beg-VERSUCH-TRY_FP: really not necessary -------------------------
% Ich wollte im Tabellentitel die Anzahl drin haben, aber \FPeval macht mit
% \rowno Probleme:
% \caption{The $ \rowno -2 $ largest primes found by the GIMPS project (as of January 2018)}
% TESTTEST: \FPeval{\result}{clip( 8 - 2 )}  Ausgabe: $8-2=\result$ \\
% TESTTEST: Value of rowno: \value{\rowno} \\
%%ERROR  \FPeval{\result}{clip(\value{\rowno} - 2)} TESTTEST: Ausgabe: $rowno-2=\result$
% \newcommand{\variableforfpevalx}{0}
% TESTTEST: Ausgabe variableforfpevalx direkt: \variableforfpevalx \\
% TESTTEST: \FPeval{\result}{clip( \variableforfpevalx - 2 )}  Ausgabe: $0-2=\result$ \\
% \newcommand{\variableforfpevaly}{\value{\rowno}}
% TESTTEST: Ausgabe variableforfpevaly direkt: \variableforfpevaly \\
%%ERROR  TESTTEST: \FPeval{\result}{clip( \variableforfpevaly - 2 )}  Ausgabe: $0-2=\result$ \\
% TESTTEST: Value of rowno: \value{\rowno} \\
%%% end-VERSUCH-TRY_FP: really not necessary -------------------------

@@

\vskip +25 pt
\paragraph*{Strong pseudo prime numbers}%
\index{prime number!strong pseudo prime}\index{number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
  In 1976 an efficient probabilistic primality test was published by Prof. Rabin,
  based on a number theoretic result of Prof. Miller from the year before.\\
  Prof. Rabin worked at the Harvard and Hebrew University.
  Michael Oser Rabin was born 1931 in Breslau. His family emigrated in
  1935 to Palestine because of the Nazi politics.
  \\
  Prof. Miller worked at the Carnegie-Mellon University, School of Computer
  Science.\\
  The functionality of the  Miller-Rabin test is explained in detail in
  \cite{Witten2010b} using a Python\index{Python} program.
}%
: It is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.
